\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-25 latex/article_vol2.Rnw
Mostly there

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and
\pkg{data.table} \citep{data.table} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules} \citep{editrules}, \pkg{validate} \citep{validate},
\pkg{DataCombine} \citep{DataCombine}, and
\pkg{janitor} \citep{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \code{clean()} on the \code{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

@@

\begin{center}
\begin{tikzpicture}[node distance = 2cm, auto,thick,scale=0.75, every node/.style={transform shape}]
    % Place nodes
    \node [block] (init) {Get next variable and run pre-checks};
    \node [cloud, above of=init] (input) {Input \texttt{data.frame} or \texttt{tibble}};
%    \node [cloud, right of=init] (system) {system};
    \node [decision, right of=init, node distance=4cm] (precheck) {Is variable suitable for inclusion};
    \node [block, right of=precheck, node distance=4cm] (summarize)
    {Run \texttt{summarize()} to produce summary table};
    \node [block, below of=summarize, node distance=3cm] (visualize)
    {Run \texttt{visualize()} to plot variable};
    \node [block, below of=visualize, node distance=2.5cm] (check)
    {Call \texttt{check()} to run error checks};
    \node [decision, below of=check, node distance=2.7cm] (done) {More
      variables?};
    \node [block, right of=done, node distance=3.5cm] (stop) {Write
      \proglang{R} markdown file};
    \node [cloud, below of=stop, node distance=3.5cm] (render) {Render
      markdown and possiby open};
    % Draw edges
    \path [line] (summarize) -- (visualize);
    \path [line] (visualize) -- (check);
    \path [line] (check) -- (done);
    \path [line] (done.south) -- +(0,-10pt) -| node [near start] {yes} (init);
%    \path [line] (identify) -- (evaluate);
%    \path [line] (evaluate) -- (decide);
%    \path [line] (init) -| node [near start] {yes} (precheck);
    \path [line] (init) -- (precheck);
    \path [line] (precheck) -- node [near start] {yes} (summarize);
    \path [line] (precheck) |- node [near start] {no} (done);
    \path [line] (done) -- node [near start] {no} (stop);
%    \path [line] (update) |- (identify);
%    \path [line] (decide) -- node {no}(stop);
    \path [line,dashed] (input) -- (init);
    \path [line] (stop) -- (render);
%    \path [line,dashed] (system) -- (init);
%    \path [line,dashed] (system) |- (evaluate);
\end{tikzpicture}
\end{center}

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \code{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using ggplot2 &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \code{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

@@

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \code{clean()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage
\item As part of the summarize/visualize/check (SVC) steps
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \code{clean()}, and above we have shown examples of how to
tweak them to modify the data cleaning outputs. However, if for
instance the dataset at hand requires new, additional checks, then more control is needed, and
Section~\ref{sec:extending} explains how to modify and expand the
possibilities by producing new summary-, visual- and check functions.

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-var2.pdf}
\end{center}
\label{fig:example2}
\caption{Output from running \code{visualize()} on the variable \texttt{var2} from the
\code{toyData} dataset.}
\end{figure}

@@

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Schunk}
\begin{Sinput}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended for]))
\end{Sinput}
\end{Schunk}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{mySummaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

@@

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures in a
\pkg{dataMaid} output document. Writing a \code{visualFunction} is slightly
more complicated than writing a \code{summaryFunction}. This follows from
the fact that \code{visualFunctions} need to be able to output standalone
code for plots in order for \code{clean()} to build standalone
\pkg{rmarkdown} files. We recommend using the following structure:
\begin{Schunk}
\begin{Sinput}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments for the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
\end{Sinput}
\end{Schunk}

@@

As with the summary function, we call \code{visualFunction()} to
register our newly created function:
%. Like before, it accepts the
%user-defined function as the first argument, and two additional
%arguments description which adds explanatory flavour text as an
%attribute of the function, and a vector which determines which classes
%the user-defined function should be applied to (which is also stored
%as an attribute).
\begin{Schunk}
\begin{Sinput}
myVisualFunction <- visualFunction(myVisualFunction,
  description = "[Some text describing the visualFunction]",
  classes = c([data types that this function is intended for]))
  )
\end{Sinput}
\end{Schunk}

@@

First, a template for the full-variable check function type, where we
first define the function and subsequently register it as a check
function using \code{checkFunction()}:
\begin{Schunk}
\begin{Sinput}
myFullVarCheckFunction <- function(v, ...) {
  [do your check]
  problem <- [is there a problem? TRUE/FALSE]
  message <- "[message describing the problem, if any]"
  checkResult(list(problem = problem,
    message = message,
    problemValues = NULL))
}

myFullVarCheckFunction <- checkFunction(myFullVarCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
  )
\end{Sinput}
\end{Schunk}

@@

If problematic values are to be identified, the template from above
should be expanded to follow a slightly more complicated structure:
\begin{Schunk}
\begin{Sinput}
myProbValCheckFunction <- function(v, nMax, maxDecimals, ...) {
	[do your check]
	problem <- [is there a problem? TRUE/FALSE]
	problemValues <- [vector of values in v that are problematic]
	problemStatus <- list(problem = problem,
	                      problemValues = problemValues)

@@

myProbValCheckFunction <- checkFunction(myProbValCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
)
\end{Sinput}
\end{Schunk}
One comment should be devoted to the helper function,
\code{messageGenerator()}.  This function's sole purpose is aiding
consistent styling of all \code{checkFunction} messages. The function
simply pastes together the \code{problemMessage} and the
\code{problemValues}, with the latter being quoted and sorted
alphabetically. If the \code{nMax} argument to \code{messageGenerator()} is
not \code{Inf}, only the first \code{nMax} problem values will be pasted
onto the message, accompanied by a comment about how many problem
values were left out (if any).  Note that printing quotes in
\pkg{rmarkdown} requires an extensive amount of character escaping, so
opting for \code{messageGenerator()} really is the easiest solution.

In the template above, the argument \code{maxDecimals} is not in use. This
argument should be used to round off the \code{problemValues} passed to
\code{messageGenerator()}, if they are numerical.  This can be done by
adding an extra line of code after defining \code{problemStatus} in the template
above:
\begin{Schunk}
\begin{Sinput}
myProbValCheckFunction <- function(v, nMax, maxDecimals, ...) {
 ...
  problemStatus <- list(problem = problem,
    problemValues = problemValues)

  if (!is.null(problemValues)) {
    problemStatus$problemValues <- round(problemValues, maxDecimals)
  }
 ...
}
\end{Sinput}
\end{Schunk}
%$
Now, problematic values will be rounded in the outputted message, while they
will still appear in their original format under the entry \code{\$problemValues} %$
in the outputted \code{checkResult}.

@@

\caption{Mosaic plot generated by the user-defined visualization
  function \texttt{mosaicVisual}.}
\label{fig:mosaicPlot}
\end{figure}

@@

\subsubsection{identifyColons --- a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We define
it using the helper function \code{messageGenerator()} to obtain a properly
escaped message, and we use \code{checkResult} to make its output print
neatly

XXX comments

@@

  # Use regular expressions to identify colons between two words
  problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
                            function(x) all(x != -1))]

@@

  problemStatus <- list(problem = problem,
                        problemValues = problemValues)
  # Use the messagegenerator function to produce the output
  # message from the problemStatus and problemMessage
  outMessage <- messageGenerator(problemStatus, problemMessage, nMax)

identifyColons <- checkFunction(identifyColons,
    description = "Identify colons surrounded by alphanumeric characters",
    classes = c("character", "factor", "labelled"))
@

@@

On the other hand, if a pdf document is to be produced, a few extra
lines need to be added to the preamble of the master \pkg{rmarkdown}
document --- otherwise, the two-column layout code will produce an
error. The following is an example of how such a master document
preamble YAML might look like and how the \code{dataMaid\_toyData.Rmd} file can
then be included:
{\small
\begin{Verbatim}
---
output: pdf_document
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---

>>>>

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and
\pkg{data.table} \citep{data.table} have made data wrangling a lot
easier, but only a handful of packages such as \pkg{editrules}
\citep{editrules}, \pkg{validate} \citep{validate}, \pkg{DataCombine}
\citep{DataCombine}, \pkg{janitor} \citep{janitor}, and
\pkg{DescTools} \citep{DescTools} attempt to implement systematic,
reproducible data cleaning.  These packages use different approaches
for data cleaning: \pkg{editrules} and \pkg{validate} provide
frameworks for setting up and checking constraints on the variables,
while \pkg{DataCombine} and \pkg{janitor} both provide a few functions
for identifying problems (e.g, duplicates, dates coded as numbers,
etc.) in data.

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{Example output from running \code{clean()} on the \code{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example1}
\end{figure}

@@

\begin{center}
\begin{tikzpicture}[node distance = 2cm, auto,thick,scale=0.75, every node/.style={transform shape}]
    % Place nodes
    \node [block] (init) {Get next variable and run pre-checks};
    \node [cloud, above of=init] (input) {Input \code{data.frame} or \code{tibble}};
%    \node [cloud, right of=init] (system) {system};
    \node [decision, right of=init, node distance=4cm] (precheck) {Is variable suitable for inclusion};
    \node [block, right of=precheck, node distance=4cm] (summarize)
    {Run \code{summarize()} to produce summary table};
    \node [block, below of=summarize, node distance=3cm] (visualize)
    {Run \code{visualize()} to plot variable};
    \node [block, below of=visualize, node distance=2.5cm] (check)
    {Call \code{check()} to run error checks};
    \node [decision, below of=check, node distance=2.7cm] (done) {More
      variables?};
    \node [block, right of=done, node distance=3.5cm] (stop) {Write
      \proglang{R} markdown file};
    \node [cloud, below of=stop, node distance=3.5cm] (render) {Render
      markdown and possiby open};
    % Draw edges
    \path [line] (summarize) -- (visualize);
    \path [line] (visualize) -- (check);
    \path [line] (check) -- (done);
    \path [line] (done.south) -- +(0,-10pt) -| node [near start] {yes} (init);
%    \path [line] (identify) -- (evaluate);
%    \path [line] (evaluate) -- (decide);
%    \path [line] (init) -| node [near start] {yes} (precheck);
    \path [line] (init) -- (precheck);
    \path [line] (precheck) -- node [near start] {yes} (summarize);
    \path [line] (precheck) |- node [near start] {no} (done);
    \path [line] (done) -- node [near start] {no} (stop);
%    \path [line] (update) |- (identify);
%    \path [line] (decide) -- node {no}(stop);
    \path [line,dashed] (input) -- (init);
    \path [line] (stop) -- (render);
%    \path [line,dashed] (system) -- (init);
%    \path [line,dashed] (system) |- (evaluate);
\end{tikzpicture}
\end{center}

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \code{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using \pkg{ggplot2} &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \code{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

@@

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \code{clean()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage.
\item As part of the summarize/visualize/check (SVC) steps.
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \code{clean()}, and above we have shown examples of how to
tweak them to modify the data cleaning outputs. However, if for
instance the dataset at hand requires new, additional checks, then more control is needed, and
Section~\ref{sec:extending} explains how to modify and expand the
possibilities by producing new summary-, visual- and check functions.

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-var2.pdf}
\end{center}
\caption{Output from running \code{visualize()} on the variable \code{var2} from the
\code{toyData} dataset.}
\label{fig:example2}
\end{figure}

@@

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended
  for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{mySummaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

@@

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures in a
\pkg{dataMaid} output document. Writing a \code{visualFunction} is slightly
more complicated than writing a \code{summaryFunction}. This follows from
the fact that \code{visualFunctions} need to be able to output standalone
code for plots in order for \code{clean()} to build standalone
\pkg{rmarkdown} files. We recommend using the following structure:
\begin{Verbatim}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments for the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
\end{Verbatim}

@@

As with the summary function, we call \code{visualFunction()} to
register our newly created function:
%. Like before, it accepts the
%user-defined function as the first argument, and two additional
%arguments description which adds explanatory flavour text as an
%attribute of the function, and a vector which determines which classes
%the user-defined function should be applied to (which is also stored
%as an attribute).
\begin{Verbatim}
myVisualFunction <- visualFunction(myVisualFunction,
  description = "[Some text describing the visualFunction]",
  classes = c([data types that this function is intended for]))
  )
\end{Verbatim}

@@

First, a template for the full-variable check function type, where we
first define the function and subsequently register it as a check
function using \code{checkFunction()}:
\begin{Verbatim}
myFullVarCheckFunction <- function(v, ...) {
  [do your check]
  problem <- [is there a problem? TRUE/FALSE]
  message <- "[message describing the problem, if any]"
  checkResult(list(problem = problem,
    message = message,
    problemValues = NULL))
}

myFullVarCheckFunction <- checkFunction(myFullVarCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
  )
\end{Verbatim}

@@

If problematic values are to be identified, the template from above
should be expanded to follow a slightly more complicated structure:

\begin{Verbatim}
myProbValCheckFunction <- function(v, nMax, maxDecimals, ...) {
	[do your check]
	problem <- [is there a problem? TRUE/FALSE]
	problemValues <- [vector of values in v that are problematic]
	problemStatus <- list(problem = problem,
	                      problemValues = problemValues)

@@

myProbValCheckFunction <- checkFunction(myProbValCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
)
\end{Verbatim}

One comment should be devoted to the helper function,
\code{messageGenerator()}.  This function's sole purpose is aiding
consistent styling of all \code{checkFunction} messages. The function
simply pastes together the \code{problemMessage} and the
\code{problemValues}, with the latter being quoted and sorted
alphabetically. If the \code{nMax} argument to \code{messageGenerator()} is
not \code{Inf}, only the first \code{nMax} problem values will be pasted
onto the message, accompanied by a comment about how many problem
values were left out (if any).  Note that printing quotes in
\pkg{rmarkdown} requires an extensive amount of character escaping, so
opting for \code{messageGenerator()} really is the easiest solution.

In the template above, the argument \code{maxDecimals} is not in use. This
argument should be used to round off the \code{problemValues} passed to
\code{messageGenerator()}, if they are numerical.  This can be done by
adding an extra line of code after defining \code{problemStatus} in the template
above:

\begin{Verbatim}
myProbValCheckFunction <- function(v, nMax, maxDecimals, ...) {
 ...
  problemStatus <- list(problem = problem,
    problemValues = problemValues)

  if (!is.null(problemValues)) {
    problemStatus$problemValues <- round(problemValues, maxDecimals)
  }
 ...
}
\end{Verbatim}
% $

Now, problematic values will be rounded in the outputted message, while they
will still appear in their original format under the entry \code{\$problemValues} %$
in the outputted \code{checkResult}.

@@

\caption{Mosaic plot generated by the user-defined visualization
  function \code{mosaicVisual()}.}
\label{fig:mosaicPlot}
\end{figure}

@@

\subsubsection{identifyColons --- a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We
define it using the helper function \code{messageGenerator()} to
obtain a properly escaped message, and we use \code{checkResult()} to
make its output print neatly. In the code we use regular expressions
through the \code{gregexpr()} function to identify colons between two
words.

@@

  problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
                            function(x) all(x != -1))]

@@

  problemStatus <- list(problem = problem,
                        problemValues = problemValues)

  checkResult(list(problem = problem,
                   message = outMessage,
                   problemValues = problemValues))
}
@

YYY en kommentar om noget her, hvad der sker nedenfor

<<>>=
identifyColons <- checkFunction(identifyColons,
    description = "Identify colons surrounded by alphanumeric characters",
    classes = c("character", "factor", "labelled"))
@

@@

On the other hand, if a pdf document is to be produced, a few extra
lines need to be added to the preamble of the master \pkg{rmarkdown}
document --- otherwise, the two-column layout code will produce an
error. The following is an example of how such a master document
preamble YAML might look like and how the \code{dataMaid\_toyData.Rmd} file can
then be included:

{\small
\begin{Verbatim}
---
output: pdf_document
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---

### 2017-01-25 latex/article_vol2.tex
Mostly there

\subsubsection{identifyColons --- a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We define
it using the helper function \code{messageGenerator()} to obtain a properly
escaped message, and we use \code{checkResult} to make its output print
neatly

XXX comments

\begin{Schunk}
\begin{Sinput}
R> identifyColons <- function(v, nMax = Inf, ... ) {
+    v <- unique(na.omit(v))
+    problemMessage <- "Note: The following values include colons:"
+    problem <- FALSE
+    problemValues <- NULL
+  
+    # Use regular expressions to identify colons between two words
+    problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
+                              function(x) all(x != -1))]
+  
+    if (length(problemValues) > 0) {
+      problem <- TRUE
+    }
+  
+    problemStatus <- list(problem = problem,
+                          problemValues = problemValues)
+    # Use the messagegenerator function to produce the output
+    # message from the problemStatus and problemMessage
+    outMessage <- messageGenerator(problemStatus, problemMessage, nMax)
+  
+    checkResult(list(problem = problem,
+                     message = outMessage,
+                     problemValues = problemValues))
+  }
R> identifyColons <- checkFunction(identifyColons,
+      description = "Identify colons surrounded by alphanumeric characters",
+      classes = c("character", "factor", "labelled"))
\end{Sinput}
\end{Schunk}

>>>>

\subsubsection{identifyColons --- a new checkFunction with problem values}
The last function mentioned above is \code{identifyColons()}. We
define it using the helper function \code{messageGenerator()} to
obtain a properly escaped message, and we use \code{checkResult()} to
make its output print neatly. In the code we use regular expressions
through the \code{gregexpr()} function to identify colons between two
words.

\begin{Schunk}
\begin{Sinput}
R> identifyColons <- function(v, nMax = Inf, ... ) {
+    v <- unique(na.omit(v))
+    problemMessage <- "Note: The following values include colons:"
+    problem <- FALSE
+    problemValues <- NULL
+  
+    problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
+                              function(x) all(x != -1))]
+  
+    if (length(problemValues) > 0) {
+      problem <- TRUE
+    }
+  
+    problemStatus <- list(problem = problem,
+                          problemValues = problemValues)
+  
+    outMessage <- messageGenerator(problemStatus, problemMessage, nMax)
+  
+    checkResult(list(problem = problem,
+                     message = outMessage,
+                     problemValues = problemValues))
+  }
\end{Sinput}
\end{Schunk}

YYY en kommentar om noget her, hvad der sker nedenfor

\begin{Schunk}
\begin{Sinput}
R> identifyColons <- checkFunction(identifyColons,
+      description = "Identify colons surrounded by alphanumeric characters",
+      classes = c("character", "factor", "labelled"))
\end{Sinput}
\end{Schunk}
